Complementing theory on adolescent emotion regulation using  machine learning

### 2022-02-22 manuscript/manuscript.Rmd
update 2022-02-22 15:25:11

<!-- TS: Emotion regulation development is a critical challenge in adolescence. -->
<!-- TS: Emotion dysregulation has severe consequences for individual adolescents, and society at large. -->
TS: Therefore, substantial research has addressed this topic.
TS: Most preceding research has been limited in scope, focusing on only a few theoretically-driven predictors.
TS: Machine learning makes it possible to cast a wider net, and explore all (available) potentially relevant predictors.
TS: Confirmatory versus exploratory research: Define the distinction.
TS: Some longitudinal research falls between these categories. For example, complex longitudinal studies, where prior research is sufficient to lead us to hypothesize *some* associations between two constructs, but not sufficiently specific to let us hypothesize which parameters will be significant.
TS: Instead, us

A key challenge in adolescence is developing mature emotion regulation abilities [@zimmermannEmotionRegulationEarly2014]. 
Adolescents experience biological, cognitive, and social changes that create an opportunity for positive emotion regulation development [@steinbergAgeOpportunityLessons2014].
Still, as many as 20% of adolescents instead experience severe emotion regulation difficulties [@leeAdolescentMentalHealth2014],
with potentially lifelong implications for mental health and well-being [@aldaoEmotionregulationStrategiesPsychopathology2010].
These difficulties in emotion regulation can cause substantial distress to the individual,
and incur a sizable societal cost.
Researchers thus seek to understand what risk factors predispose some adolescents to emotion dysregulation,
so that effective screening and interventions can be developed.
Despite abundant empirical research, however,
an overarching understanding is still lacking because
the literature is fragmented across subdisciplines 
and lacks consistent terminology and guiding theory [@vanlissaMappingPhenomenaRelevant2021].

The present study seeks to gain a more comprehensive understanding of adolescent emotion regulation development by using machine learning methods for rigorous exploration.
<!-- These predictors may include both risk factors that render some youth susceptible to emotion regulation difficulties, and correlates that can be used to identify at-risk youth. -->
The SEM-forests algorithm was used to
identify the most important early indicators of developmental trajectories
of emotion regulation across adolescence, from a large set of potential candidates.
The resulting insights can guide future research and interventions to focus on the most important predictors,
and lays the foundation for formation and revision of overarching theory, such that essential phenomena are adequately represented.

Recently, a text mining systematic review of 6305 publications on adolescents' emotion regulation has revealed what phenomena

@@

```{r networks1, eval = TRUE, results='asis', out.width = "90%"}
knitr::include_graphics("tmnetworks_small.png")
```

@@

```{r, out.height="50%"}
knitr::include_graphics("groups.png")
```

@@

```{r}
knitr::include_graphics("tree.png")
```

@@

```{r, out.height="100%"}
vim <- readRDS("varimp_comb_21-08-2021.RData")
vim
```

@@

```{r, out.width="40%"}
knitr::include_graphics("plots/pdp1-12.png")
```

@@

```{r, out.width="40%"}
knitr::include_graphics("plots/pdp13-24.png")
```

@@

```{r, out.width="40%"}
knitr::include_graphics("plots/pdp25-36.png")
```

@@

```{r, out.width="40%"}
knitr::include_graphics("plots/pdp37-44.png")
```

@@

## {.fullslide}

![](polarplot2.png)

>>>>

<!-- TS: Emotion regulation development is a critical challenge in adolescence. -->
<!-- TS: Emotion dysregulation has severe consequences for individual adolescents, and society at large. -->
<!-- TS: Therefore, substantial research has addressed this topic. -->
<!-- TS: Most preceding research has been limited in scope, focusing on only a few theoretically-driven predictors. -->
<!-- TS: Machine learning makes it possible to cast a wider net, and explore all (available) potentially relevant predictors. -->
<!-- TS: Confirmatory versus exploratory research: Define the distinction. -->
<!-- TS: Some longitudinal research falls between these categories. For example, complex longitudinal studies, where prior research is sufficient to lead us to hypothesize *some* associations between two constructs, but not sufficiently specific to let us hypothesize which parameters will be significant. -->
<!-- TS: Instead, us -->

A key challenge in adolescence is developing mature emotion regulation abilities [@zimmermannEmotionRegulationEarly2014]. 
Adolescents experience biological, cognitive, and social changes that create an opportunity for positive emotion regulation development [@steinbergAgeOpportunityLessons2014].
Still, as many as 20% of adolescents instead experience severe emotion regulation difficulties [@leeAdolescentMentalHealth2014],
with potentially lifelong implications for mental health and well-being [@aldaoEmotionregulationStrategiesPsychopathology2010].
These difficulties in emotion regulation can cause substantial distress to the individual,
and incur a sizable societal cost.
A comprehensive understanding of the risk factors that predispose some adolescents to emotion dysregulation could aid in the development of effective screening and interventions.
However, despite an abundance of empirical research,
many studies are narrow in scope, considering only a few predictors,
the literature is fragmented across subdisciplines, and lacks a unifying theoretical framework
<!-- that is both specific to adolescence and comprehensive in scope  -->
[@vanlissaMappingPhenomenaRelevant2021].

The present study seeks to overcome these limitations
by using data-driven insights to complement and reflect on existing theory.
To this end, a broad range of potential risk factors were analyzed using machine learning,
and ranked in terms of predictive importance.
<!-- These predictors may include both risk factors that render some youth susceptible to emotion regulation difficulties, and correlates that can be used to identify at-risk youth. -->
Specifically,
the SEM-forests algorithm was used to
identify the most important early indicators of developmental trajectories
of emotion regulation across adolescence.
The resulting insights indicate which factors are most predictive of adolescents' diverging destinies in terms of emotion regulation development.
Armed with this knowledge, future research and interventions can focus on the most important predictors.
Moreover, these insights can guide the formation of overarching theory that adequately represents essential phenomena.
<!-- complement existing theory on adolescent emotion regulation using data-driven insights from machine learning. -->

##

## Emotion dysregulation in adolescence

Emotion regulation,
defined as the ability to manage emotional experience and expression,
is critical for mental health4–6 and social competence2 throughout life. There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions1. 
According to Crone and Dahl's neurodevelopmental model1,
staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence:
A period wherein the emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind8,21,22.
This imbalance motivates most adolescents to engage in adaptive exploration:
Seeking out experiences required to develop mature socio-emotional competences.
At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation1,10,
<!-- exposes youth to a risk of emotion dysregulation, -->
for example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
or persevere in pursuing adaptive goals when facing obstacles.
Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions2,9,11,14,
which has been linked to the emergence of several mental disorders5,10.
<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
<!-- but renders some of them vulnerable to emotion dysregulation1,10. Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11. -->
<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->

Empirical support for this notion of diverging destinies11 is evident from studies that examine developmental trajectories of emotion regulation in adolescence: 
Such studies consistently reveal large between-individual heterogeneity 2,7,9.
An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
This involves three important challenges:
The first is that the number of potentially relevant risk factors is large,
the second is that there is a paucity of relevant theory to narrow down this number,
and the third is that methods commonly used in the field are not well suited for large numbers of predictors.
We argue that machine learning methods that perform variable selection overcome these challenges.

<!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards “person-centered” research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23. Seminal studies have applied “clustering” methods to identify youth with similar developmental trajectories14,24–26. Adolescents may follow similar trajectories for different reasons, however, and person-centered methods are too complex to accommodate more than a handful of theoretically-derived predictors. Unfortunately, relevant theory on adolescent emotion regulation development is limited27–29. The field is therefore at an impasse: We know that some adolescents are more susceptible to emotion dysregulation than others, but lack tools and theory to identify important predictors of individual development30. -->

## Potential predictors

<!-- Maybe review theory briefly? -->
A recent literature review set out to identify relevant phenomena in relation to adolescents' emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
It used a new technique, text mining systematic review,
to identify relevant phenomena based on frequently occurring terms in the author keywords and abstracts of 6305 publications.
Whereas narrative reviews are typically restricted to small corpora and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008],
text mining systematic reviews are more comprehensive and derive insights from the literature through a relatively transparent and objective procedure - at the expense of reflection and meaning-making.

The findings of that systematic review indicated that internalizing and externalizing problems and mental health-related outcomes were most frequently studied in conjunction with emotion regulation. 
Other relevant phenomena reflected themes discussed in prior narrative reviews,
including cognitive development, parenting, and external stressors [cf. @coe-odessEmergentEmotionsAdolescence2019].
Other relevant phenomena were underrepresented in prior theory and reviews,
such as substance use, structural disadvantage,
identity, morality, sex and sexual development.
This relatively comprehensive overview offers a starting point for empirical investigations of predictors of adolescents' emotion regulation development.
<!-- To this end, the present study sought to identify operationalizations of these relevant phenomena in a large-scale panel study. -->

## Complementing theory using machine learning

As mentioned before, the field of adolescent emotion regulation is prolific, but lacks overarching theory.
For guidance on how to remedy this shortcoming, we turn to philosophy of science.
De Groot’s "empirical cycle" constitutes a model of the production of knowledge through scientific research [@degrootMethodologieGrondslagenVan1961].
It consists of five phases: *observation* of phenomena to be explained; *induction* of hypotheses; *deduction* or derivation of testable predictions; *testing* these predictions in new data; and finally *evaluation*: reflecting back on the theories and hypotheses that gave rise to the tested predictions.
The model is cyclical because *evaluation* offers new interpretative data for the researcher to observe - such as rejected hypotheses and serendipitous findings.

<!-- The empirical cycle is merely one model, -->
<!-- and there is a pluriformity of other epistemological approaches to knowledge production. -->
<!-- Nonetheless, u -->
Unsubstantiated or undisclosed deviations from the empirical cycle may signal practices that limit scientific progress.
Consider three examples:
First, conducting many analyses and then writing a paper that presents significant findings as tests of a-priori hypotheses, or "hypothesizing after the results are known", constitutes a questionable research practice that leads to false-positive results [@wichertsDegreesFreedomPlanning2016].
Second, failure to engage in evaluation and reflect back on theory in light of empirical findings leads to flexible, "invulnerable" theories that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
Third, and particularly relevant to the present study,
is failure to complete the cycle.
In a productive field of study, the empirical cycle is regularly completed;
evaluation of theories in the light of empirical findings allows scholars to hone those theories,
which in turn yield increasingly precise predictions.
We argue that, the paucity of overarching theory in our field can be remedied by completing the empirical cycle.

The phases of the empirical cycle map onto two modes of conducting research:
Confirmatory research, which involves the phases of deduction, testing, and evaluation, and exploratory research, which involves the phases of observation and induction  [@wagenmakersCreativityVerificationCyclePsychological2018].
Although developmental psychology is dominated by quantitative positivism,
relatively few studies are explicitly exploratory [@demuthQualitativeMethodologyDevelopmental2015].

and devoting greater attention to exploratory research and theory formation.

<!-- In this model, the hypothesis plays an important role. -->
<!-- It is a testable prediction that can be supported or falsified; depending on the results, we reflect on old knowledge – which includes theory – and make necessary adjustments that lead to new hypotheses. -->
<!-- A hypothesis can thus be a specific prediction that delivers new knowledge when it is false; -->
<!-- this drives a scientific field forward. -->

Prior theory and empirical research thus suggests many phenomena potentially related to heterogeneity in adolescents' emotion regulation development.
What is missing is an understanding of the relative utility of each of these phenomena in predicting developmental outcomes.
Individual studies do not suffice for this purpose as
they tend to sample from different populations,
use different research designs,
different control variables and statistical models.
Relatedly, many studies tend to examine only a handful of predictors,
often guided by a theory with narrow scope - 
, it is hard to tell how important the variables are, relative to other variables considered by other theories (or variables not considered at all). Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest

  Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest.

@@

@@

@@

@@

@@

@@

@@

@@

@@
